\section{Related Work}

In this section, we provide a comprehensive review of related research in simultaneous localization and mapping (SLAM), with a focus on three key areas: (1) traditional SLAM algorithms and their evolution, (2) IMU-visual sensor fusion approaches, and (3) brain-inspired SLAM systems. These three research directions represent the foundation upon which our proposed NeuroNav system, a HART+Transformer enhanced NeuroSLAM variant, is built.

\subsection{Traditional SLAM Algorithms}

\subsubsection{Visual SLAM Systems}

The SLAM problem has been extensively studied over the past three decades, with significant advances in both theoretical foundations and practical implementations \cite{Cadena2016}. Traditional SLAM systems can be broadly categorized based on their sensor modalities and algorithmic approaches.

\textbf{Monocular Visual SLAM.} 
Monocular visual SLAM has been a popular research direction due to the low cost and simplicity of camera sensors. 
ORB-SLAM \cite{MurArtal2015, MurArtal2017} represents a milestone in feature-based visual SLAM, utilizing oriented FAST and rotated BRIEF (ORB) features for robust tracking, mapping, and loop closure detection. 
The system employs a bag-of-words approach for place recognition and achieves real-time performance on standard CPUs. 
However, ORB-SLAM struggles with scale ambiguity inherent to monocular systems and requires substantial computational resources for feature extraction and matching.

PTAM (Parallel Tracking and Mapping) \cite{Klein2007} pioneered the separation of tracking and mapping into parallel threads, enabling real-time performance on small-scale environments. 
While groundbreaking, PTAM is limited to small workspaces and lacks robust loop closure detection.
LSD-SLAM \cite{Engel2014} took a different approach by using direct image alignment instead of features, achieving semi-dense reconstruction. 
However, it is sensitive to photometric changes and requires careful parameter tuning.

More recently, direct sparse odometry (DSO) \cite{Engel2018} improved upon LSD-SLAM by incorporating photometric calibration and optimizing a joint photometric and geometric error. 
DSO demonstrates superior accuracy compared to feature-based methods in well-textured environments but struggles in low-texture scenarios.
SVO (Semi-direct Visual Odometry) \cite{Forster2014, Forster2017} combines the advantages of feature-based and direct methods, achieving high-speed performance by minimizing reprojection error at sparse feature locations.

\textbf{RGB-D and Stereo SLAM.}
RGB-D cameras have enabled dense 3D reconstruction by providing direct depth measurements.
KinectFusion \cite{Newcombe2011} demonstrated real-time dense surface reconstruction using truncated signed distance function (TSDF) representation, though it was limited to small-scale environments.
ElasticFusion \cite{Whelan2016} extended this work to enable large-scale dense mapping with a surfel-based representation and loop closure detection.

Stereo SLAM systems, such as ORB-SLAM2 \cite{MurArtal2017}, leverage baseline information to recover absolute scale while maintaining the flexibility of passive sensing.
However, stereo systems require precise calibration and struggle in textureless regions where feature matching is unreliable.

\textbf{Semantic SLAM.}
Recent advances have incorporated semantic understanding into SLAM systems.
SLAM++ \cite{Salas2013} introduced object-level SLAM by detecting and tracking known objects in the environment, enabling more compact and meaningful map representations.
SemanticFusion \cite{McCormac2017} combines dense reconstruction with CNN-based semantic segmentation, creating semantically labeled 3D maps.
While semantic SLAM provides richer environmental understanding, it relies heavily on pre-trained object detection models and may fail in novel or cluttered environments.

\subsubsection{Loop Closure and Place Recognition}

Loop closure detection is critical for correcting accumulated drift in long-term SLAM.
FAB-MAP (Fast Appearance-Based Mapping) \cite{Cummins2008, Paul2010} pioneered probabilistic place recognition using a bag-of-words model with a Chow-Liu tree to capture feature co-occurrence.
SeqSLAM \cite{Milford2012} demonstrated robust place recognition under extreme appearance changes by matching sequences of images rather than individual frames, though it requires storing entire image sequences.

More recent deep learning approaches, such as NetVLAD \cite{Arandjelovic2016}, learn compact image representations for place recognition, achieving state-of-the-art performance.
However, these methods typically require large-scale training datasets and may not generalize well to novel environments.

\subsubsection{Limitations of Traditional SLAM}

Despite significant progress, traditional SLAM systems face several fundamental challenges:
(1) \textit{Computational complexity}: Many state-of-the-art systems require substantial computational resources for feature extraction, matching, and optimization, limiting their deployment on resource-constrained platforms.
(2) \textit{Environmental assumptions}: Most systems assume static, well-lit environments with sufficient texture, failing in dynamic or challenging conditions.
(3) \textit{Scalability}: Long-term operation in large-scale environments leads to increasing computational burden and memory requirements.
(4) \textit{Robustness}: Traditional systems often suffer from catastrophic failures due to perceptual aliasing, rapid motion, or sensor degradation.


\subsection{IMU-Visual Sensor Fusion}

Integrating inertial measurement units (IMUs) with visual sensors has emerged as a powerful approach to address limitations of pure visual SLAM, providing complementary information for robust state estimation.

\subsubsection{Tightly-Coupled Visual-Inertial Odometry}

Tightly-coupled approaches fuse visual and inertial measurements within a unified optimization framework, leveraging the complementary nature of these modalities.

MSCKF (Multi-State Constraint Kalman Filter) \cite{Mourikis2007} pioneered the use of EKF-based visual-inertial fusion by marginalizing feature observations to maintain computational efficiency.
However, MSCKF's reliance on a linearized system model can lead to inconsistency in highly nonlinear scenarios.

OKVIS (Open Keyframe-based Visual-Inertial SLAM) \cite{Leutenegger2015} employs nonlinear optimization over a bounded window of recent keyframes and IMU measurements, achieving accurate state estimation.
While OKVIS demonstrates superior accuracy compared to filtering approaches, its keyframe-based marginalization can discard valuable information.

ROVIO (Robust Visual Inertial Odometry) \cite{Bloesch2015} takes a direct approach by tracking image patches rather than point features, making it robust to feature-poor environments.
However, ROVIO does not perform loop closure and thus accumulates drift over long trajectories.

VINS-Mono \cite{Qin2018} represents a significant advancement in monocular visual-inertial fusion by incorporating robust initialization, loop closure via a pose graph, and four-degree-of-freedom optimization.
The system achieves state-of-the-art accuracy and has been widely adopted in both research and commercial applications.
VINS-Fusion \cite{Qin2019} extends this work to support stereo cameras and stereo+IMU configurations, demonstrating improved robustness and scale observability.

VI-ORB-SLAM \cite{MurArtal2017} integrates IMU measurements into the ORB-SLAM framework, enabling robust initialization and improved tracking performance.
However, its tight coupling increases system complexity and computational requirements.

\subsubsection{Loosely-Coupled Approaches}

Loosely-coupled methods treat visual and inertial odometry as separate modules, fusing their outputs at the state estimation level.
While computationally efficient, these approaches do not fully exploit the synergy between modalities and are more susceptible to sensor failures.

S-MSCKF \cite{Sun2018} extends MSCKF with stereo cameras, demonstrating improved performance in GPS-denied environments.
However, the increased number of measurements significantly increases computational complexity.

\subsubsection{Challenges in Visual-Inertial Fusion}

Despite significant progress, visual-inertial SLAM faces several persistent challenges:
(1) \textit{Initialization}: Accurate estimation of initial states, including IMU biases and velocity, is crucial but difficult in practice.
(2) \textit{Calibration}: Precise spatial and temporal calibration between camera and IMU is essential for tight coupling.
(3) \textit{Degeneracy}: Certain motion patterns (e.g., pure rotation) lead to unobservable states.
(4) \textit{Computational cost}: Real-time operation on embedded platforms remains challenging, especially for multi-camera systems.

Our proposed system addresses these challenges by incorporating bio-inspired neural mechanisms that provide intrinsic robustness and efficiency.


\subsection{Brain-Inspired SLAM and Neural Navigation}

Biological navigation systems demonstrate remarkable capabilities that inspire novel approaches to robot navigation and SLAM.
Animals navigate effectively in complex, dynamic environments with limited computational resources, suggesting fundamentally different strategies from traditional engineering approaches.

\subsubsection{Rodent-Inspired Navigation Systems}

RatSLAM \cite{Milford2004, Milford2008, Milford2010} pioneered the application of rodent navigation principles to robot SLAM.
The system models place cells, head direction cells, and grid cells using continuous attractor neural networks (CANNs), creating a topological-metric hybrid representation.
RatSLAM successfully demonstrated suburb-scale mapping and long-term navigation, achieving robustness to appearance changes that challenge traditional visual SLAM.
However, the original RatSLAM operates primarily in 2D environments and uses hand-crafted visual features that limit performance in complex scenarios.

Extensions of RatSLAM include BatSLAM \cite{Steckel2013}, which incorporates sonar sensing for navigation in darkness, and various improvements to the visual processing pipeline \cite{Milford2011a, Milford2011b}.

\subsubsection{3D Navigation and Spatial Representation}

Recent neuroscientific discoveries of 3D grid cells and 3D head direction cells in bats \cite{Finkelstein2015, Finkelstein2016} and humans \cite{Kim2017, Kim2018a, Kim2018b, Kim2019} have inspired efforts to extend bio-inspired SLAM to three-dimensional environments.

3D place cells represent specific 3D locations \cite{Yartsev2013, Wohlgemuth2018}, forming a sparse metric map.
3D head direction cells encode the animal's orientation in 3D space through combinations of azimuth and pitch \cite{Finkelstein2015, Page2018, Shinder2019}.
3D grid cells exhibit periodic firing patterns in volumetric space, potentially providing a universal metric for 3D path integration \cite{Finkelstein2016, Casali2019}.

Mathematical models of these 3D spatial cells have been proposed \cite{Jeffery2013, Jeffery2015}, including face-centered cubic (FCC) and hexagonal close-packed (HCP) lattice organizations for 3D grid cells.
Soman et al. \cite{Soman2018} developed a hierarchical network model of 3D spatial cells, demonstrating the computational principles underlying volumetric navigation.

Silveira et al. \cite{Silveira2013, Silveira2015} extended RatSLAM to underwater 3D environments using a 3D place cell model, though without explicit directional representation.
Yu et al. proposed NeuroSLAM for 3D environments by modeling 3D grid cells and multi-layered head direction cells, demonstrating 4-DOF pose estimation.

\subsubsection{Deep Learning and Neuromorphic Approaches}

Recent work has explored combining bio-inspired principles with modern deep learning.
Kreiser et al. \cite{Kreiser2018a, Kreiser2018b} developed spiking neural network (SNN) implementations of spatial cells running on neuromorphic hardware, achieving energy-efficient place recognition.

Attention mechanisms, inspired by selective visual processing in the brain, have shown promise in visual place recognition \cite{Chen2017}.
However, these approaches typically focus on individual components (e.g., place recognition) rather than complete SLAM systems.

\subsubsection{Limitations and Opportunities}

While brain-inspired SLAM systems demonstrate unique advantages in robustness and long-term stability, several challenges remain:
(1) \textit{Feature representation}: Most bio-inspired systems use hand-crafted features that limit performance in complex environments.
(2) \textit{3D navigation}: Few systems successfully operate in true 3D environments with complete 6-DOF pose estimation.
(3) \textit{Scalability}: Computational complexity of neural network models can limit real-time performance.
(4) \textit{Integration}: Combining bio-inspired principles with modern sensor fusion and deep learning remains largely unexplored.

Our work addresses these limitations by incorporating the HART visual attention mechanism with Transformer-based feature processing, creating a biologically-inspired yet computationally efficient approach to visual feature extraction for neural SLAM systems.


\subsection{Mapping and Localization Paradigms}

The representation of spatial information is fundamental to SLAM, with different paradigms offering distinct advantages and limitations.

\subsubsection{Metric Mapping}

Metric maps represent the environment with precise geometric information, typically as point clouds, voxel grids, or mesh structures.
Dense methods like KinectFusion \cite{Newcombe2011} and ElasticFusion \cite{Whelan2016} create detailed 3D reconstructions suitable for robot manipulation and human-robot interaction.
Sparse methods like ORB-SLAM \cite{MurArtal2015} maintain only landmark positions, reducing memory requirements while sacrificing environmental completeness.

OctoMap \cite{Hornung2013} provides a probabilistic volumetric representation using an octree structure, enabling efficient storage and uncertainty modeling.
However, metric maps scale poorly to large environments and are sensitive to sensor noise.

\subsubsection{Topological Mapping}

Topological maps represent environments as graphs of discrete locations and connections, abstracting away precise geometric details.
This representation naturally supports hierarchical planning and is robust to perceptual changes \cite{Lowry2016}.

Vision-based topological mapping systems like FAB-MAP \cite{Cummins2008} and SeqSLAM \cite{Milford2012} create place-based graph representations.
These systems excel in appearance-based loop closure detection but provide limited metric information for precise navigation.

\subsubsection{Hybrid Representations}

Hybrid approaches combine the benefits of metric and topological representations.
The experience map in RatSLAM \cite{Milford2008} creates a topological graph where each node stores local metric pose information, balancing computational efficiency with metric accuracy.

Semantic mapping approaches \cite{Salas2013, McCormac2017} create hierarchical representations with object-level topology and instance-level geometry.
These representations are compact and meaningful but rely on object detection, limiting applicability in unstructured environments.

\subsubsection{Neural Spatial Representations}

Grid cells in the mammalian brain provide a neural code for space that combines properties of metric and topological representations \cite{Hafting2005, Moser2017}.
The periodic nature of grid cell activity creates a multi-scale metric representation, while the discrete attractor states in grid cell networks enable topological robustness.

Our system leverages this neural spatial representation through continuous attractor networks, creating an inherently hybrid mapping approach that combines the precision of metric maps with the robustness of topological representations.


\subsection{Research Gap and Contributions}

Despite extensive research in traditional SLAM, visual-inertial fusion, and bio-inspired navigation, several key challenges remain unaddressed:

\begin{enumerate}
    \item \textbf{Feature representation in bio-inspired navigation}: Existing brain-inspired systems predominantly use hand-crafted visual features (e.g., intensity scanlines, whole-image matching) that limit performance in complex real-world scenarios. There is a lack of integration between modern deep visual features and bio-inspired spatial representations such as grid-cell and head-direction-cell attractor networks.
    
    \item \textbf{Attention and dual-stream visual processing}: Biological vision employs dorsal/ventral dual streams and sophisticated attention mechanisms to focus on behaviourally relevant information. In contrast, most engineering systems process all pixels uniformly or rely on a single visual pipeline, leading to computational inefficiency and vulnerability to clutter and illumination changes.
    
    \item \textbf{Long-range temporal dependencies for inertial-visual fusion}: Spatial navigation inherently involves dependencies across long sequences of multi-modal observations. However, many existing visual-inertial systems exploit only short temporal windows and do not explicitly model long-range correlations that are crucial for drift suppression and robust loop re-localisation.
    
    \item \textbf{3D bio-inspired navigation with modern sensors}: While 3D spatial cells have been modeled and implemented in NeuroSLAM-like architectures, practical deployments that tightly fuse these neural representations with modern IMU-visual front-ends in realistic 3D environments are still limited.
    
    \item \textbf{End-to-end interpretation of component contributions}: There is a lack of systematic studies that connect cortical-style visual processing, inertial sensing, and hippocampal/entorhinal representations (experience map, grid cells, head-direction cells) within a single, analysable navigation framework.
\end{enumerate}

This paper addresses these gaps by proposing a \emph{brain-inspired inertial-visual fusion system} built on the NeuroSLAM spatial representation backbone. Concretely, we integrate:
(1) a biologically-inspired dual-stream visual encoder (dorsal and ventral pathways) with HART-like attention for selectively processing salient motion and appearance cues,
(2) Transformer and LSTM-based temporal modules for capturing long-range dependencies in visual and inertial sequences,
(3) fusion of timestamp-aligned IMU measurements and visual odometry to provide vestibular-style self-motion signals to the 3D grid-cell and multilayer head-direction-cell networks, and
(4) comprehensive evaluation on both simulated CARLA towns and real EuRoC sequences to quantify the benefits of each component via ablation.
Our approach bridges the gap between bio-inspired navigation principles and modern deep inertial-visual fusion, yielding an interpretable brain-inspired navigation framework for challenging 3D environments.
