

\documentclass[11pt,journal,onecolumn,draft]{IEEEtran}


% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}
\usepackage{graphicx}
\usepackage[table,xcdraw]{xcolor}
\usepackage[TABBOTCAP]{subfigure}
\usepackage{bm}
\usepackage{upgreek} 
\usepackage{amsmath}
\usepackage{breqn}
\usepackage{color}
\usepackage{cite}
\usepackage[none]{hyphenat}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{caption2}
\usepackage{setspace}

\begin{document}

\begin{center}
	\textbf{\LARGE Cover Letter} 
\end{center}
\vspace{8pt} 
Dear Editor and Reviewers:

\hspace{1em} We want to thank you and the review team for  our submission and for the opportunity to consider the paper to Knowledge-Based Systems.  
No conflict of interest exits in the submission of this manuscript, and manuscript is approved by all authors for publication.  
I would like to declare on behalf of my co-authors that the work described was original research. 
All the authors listed have approved the manuscript that is enclosed. 

In this work, we has some contributions.
First, we propose a brain-inspired driving (BID) model and utilized brain-inspired technology to achieve more efficient and robust environmental perception and decision. It advances interpretability and robustness in autonomous driving.
Second, the experimental results show that the performance has been significantly improved across various autonomous driving tasks and achieved the end-to-end autopilot successfully. 
In addition, it offers fancy insights and methodologies for further advancing autonomous driving technology.
Therefore, we hope our manuscript can be considered for publication. 

\hspace{1em} We deeply appreciate your consideration of our manuscript.
%and we look forward to receiving comments from the reviewers. 
If you have any queries, please don’t hesitate to contact me at the address below.

\hspace{1em} Thank you and best regards.

\hspace{1em} Yours sincerely,

\hspace{1em} Haidong Wang
\vspace{20pt} 

Corresponding author:

Name: Haidong Wang

E-mail: whd@hutb.edu.cn 

\end{document}


